Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences

### 2017-09-11 paper-amia-summit-2018/amia_summit_2018.tex
update discussion section

\section*{Discussion}
By analyzing the experimental results of different communication sequence outcome prediction schemes proposed in this paper, we arrived at the following conclusions. First, using higher-order models results in better prediction accuracy compared to lower-order models, since higher-order models utilize larger context and can better capture different nuances of patient-provider communication. On the other hand, the number of states in higher-order Markov models grows exponentially with their order. Therefore, accurate estimation of transition probabilities requires much larger training set. Using smaller datasets will result in a sparsity problem, when many transitions are either not observed in the training set at all or observed only a few times, leading to missing or potentially inaccurate probability estimates. Obtaining large training sets cannot be easily accomplished in many domains, including motivational interviewing. In this study, we found out that using second-order Markov models is a reasonable trade-off between efficiency and accuracy.

Second, the overall predictive performance of HMM is substantially better than MC for both types of sequences. In particular, HMM-based method achieves near-human accuracy for predicting the success of motivational interviews. This indicates that hidden states in HMM are able to capture the structure of discourse in motivational interviews by grouping together the codes that correspond to different cognitive states of adolescents during motivational interviews, which reflect the overall progression of the interviews. This allows to reduce the dimensionality of codes in PPC sequences and consequently improve both precision and recall of the prediction method.

Third, converting natural PPC communication sequences to alternating ones results in better performance for all configurations of the proposed method. This indicates that alternating sequences emphasize the dependencies between the pairs of patient and provider codes, which results in more accurate estimates of the corresponding conditional probabilities. Better estimates of parameters in Markov models of successful and unsuccessful interviews are propagated to the next step, where they are aggregated into predictions for the entire sequence. This allows to achieve a dramatic improvement in the prediction accuracy of the method.

Fourth, the proposed method can be used to identify the most effective communication strategies at eliciting a particular type of behavioral response. Awareness of these strategies by researchers can significantly decrease the time and effort required to develop effective interventions to address many public health conditions, such as childhood obesity, and tailor these interventions to particular patient cohorts. Awareness of these strategies by the counselors can lead to greater success rate of motivational interviews.

>>>>

\section*{Discussion}
By analyzing the experimental results of different communication sequence outcome prediction schemes proposed in this paper, we arrived at the following conclusions. First, the overall predictive performance of RNN models is substantially better than probabilistic models. In particular, RNN-based method achieves near-human accuracy for predicting the success of motivational interviews. This indicates that RNN are able to capture the structure of discourse in motivational interviews by preserving long-term dependencies among the behavior codes, which reflect the overall progression of the interviews. In addition to that word embedding allows to reduce the dimensionality of codes in PPC sequences and consequently improve both precision and recall of the prediction method.

Second, computing loss with target replication at each time step results in better performance for all configurations of the proposed RNN method. This indicates that mean of the losses over all steps emphasize the dependencies between the pairs of patient and provider codes, which results in more accurate estimates of the model parameters. Better estimates of parameters in RNN models of motivational interviews are propagated to the next step based on the relative importance of intermediate output, where they are aggregated into predictions for the entire sequence. This allows to achieve an improvement in the prediction accuracy of the method.

Third, using first-order Markov model results in better prediction accuracy compared to higher-order Markov models. Because the number of states in higher-order Markov models grows exponentially with their order. Therefore, accurate estimation of transition probabilities requires much larger training set. Using smaller datasets such as under-sampled dataset will result in a sparsity problem, when many transitions are either not observed in the training set at all or observed only a few times, leading to missing or potentially inaccurate probability estimates. Obtaining large training sets cannot be easily accomplished in many domains, including motivational interviewing. In this study, we found out that using first-order Markov models is a reasonable trade-off between efficiency and accuracy.

Fourth, similar to traditional markov model, HMM achieve a dramatic improvement in the prediction accuracy when larger training set is used. This indicates that sufficient training data is required to estimate optimal hyperparameters such as number of hidden states, intitial state distribution, transition probabilities, and emission probabilities.

Fifth, the proposed method can be used to identify the most effective communication strategies at eliciting a particular type of behavioral response. Awareness of these strategies by researchers can significantly decrease the time and effort required to develop effective interventions to address many public health conditions, such as childhood obesity, and tailor these interventions to particular patient cohorts. Awareness of these strategies by the counselors can lead to greater success rate of motivational interviews.
